Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-16 manuscript.tex
edits to methods

%TO DO
%Fix figures and tables
%add usepackage for tables
%check citations in last paragraph for ter1
%reformat for AJB
%preprint somewhere
%finish fixing paragraph that jeff doesn't like
%check map details
%make figs smaller so they don't go to end of document
%Code for SAS and command line stuff on Git
%Phenotyping and genotyping data on Git (it's on Figshare?)

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE).  To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the \emph{Hopscotch} insertion may be adaptive under certain environmental conditions, we looked for significant associations between the \emph{Hopscotch} frequency and environmental variables using BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\citep{Coop2010}. We used genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional SNP (\ref{TableS3BayEnv}). Environmental data were obtained from \url{www.worldclim.org}, the Harmonized World Soil Database and \url{www.harvestchoice.org}, and summarized by principle component analysis \citep{Pyhajarvi2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \citep{Thornton2003} to calculate pairwise F\textsubscript{ST} between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator $\hat\theta_W$; pairwise nucleotide diversity $\hat\theta_\pi$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by \emph{Hopscotch} genotype we used Phylip (\url{http://evolution.genetics.washington.edu/phylip.html}) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the \emph{Hopscotch} insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \citep{Chia2012}).

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and the same phased 55K SNP data from \citep{Pyhajarvi2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map (\citep{Gerke2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in. \jri{i'd like this perl script on github, maybe in this repo or as a gist. also structure input file too. all the stuff we'd need to redo this.} \lev{See above note..not all of this or the BayEnv was script'ified} \jri{sure, even command line info should be included where possible.  idea is to maximize reproducibility -- either other people or subsequent students. for example, matt has a student who wants to work on tb1 in natural pops, and she might want to try/redo some of these analyses with the same or new data.} \lev{okay so I should just put my command line stuff with good commenting as to what is what in repository?}

\subsection*{Phenotyping of \emph{Zea mays.} ssp. \emph{parviglumis}}

To investigate the phenotypic effects of the \emph{Hopscotch} insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the \emph{Hopscotch} is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the \emph{Hopscotch} in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Following germination, 206 successful germinations were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer. \lev{it ended up being a combination of drip and hand watering because they dried out so much and did better when they had water on the leaves as well as in the soil}

To investigate the phenotypic effects of the \emph{Hopscotch} insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the \emph{Hopscotch} is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the \emph{Hopscotch} in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Successful germinations (206 individuals) were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day.

Starting on day 15, we measured tillering index, the ratio of the sum of tiller lengths to the height of the plant \citep{Briggs2007}. Tillering index has been shown to be the most effective way to observe the phenotypic effects of the \emph{Hopscotch} insertion on plant architecture in maize \citep{Clark2006}. Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index, or variation at \emph{tb1} (e.g. \emph{Hopscotch} genotype). Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (http://www.ars.usda.gov). We genotyped individuals for the \emph{Hopscotch} insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of phenotyping 1, indicating that a minimum of 71 individuals in each genotype class are needed to detect the observed effect of the \emph{Hopscotch} on tillering index. \jri{do you still have these posthoc calculations?} \lev{I believe I do, otherwise they are likely in a lab meeting slide on Dropbox -- do you want them in here?} \jri{would be good to include in a github, again so we could go back and reassess how we do things in case, for example, we decide to give the greenhouse experiment a 3rd try (yes, i'm a masochist)} \lev{I would do it a 3rd time, I'm convinced something went wrong....third time with expression analyses :) }

We performed a second phenotyping experiment (phenotyping 2) in which we germinated 372 seeds of \emph{parviglumis}, choosing equally between sites previously determined to have or not have the \emph{Hopscotch} insertion. Seeds were germinated and planted on day 7 post fruit-case removal into 2 gallon pots. Plants were watered twice daily, alternating between fertilized and non-fertilized water. We began phenotyping successful germinations (302) for tillering index on day 15 post fruit case removal, and phenotyped every five days until day 50. At day 50 we measured culm diameter between the third and fourth nodes. We extracted DNA and genotyped plants following the same guidelines as in phenotyping 1.

Resulting tillering index data for each genotype class did not meet the criteria for a repeated measures ANOVA, so we transformed the data using a Box-Cox transformation ($\alpha=0$) implemented in the car package in R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between \emph{Hopscotch} genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. \jri{please add SAS scripts/code to a gist or something. }

@@

To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion from our initial genotyping. We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. We found the \emph{Hopscotch} segregating at a frequency of 0.65 with no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find a weak positive correlation between presence of the \emph{Hopscotch} and tillering index on day 40 (p=0.0848), but no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and \emph{Hopscotch} genotype, or culm diameter and \emph{Hopscotch} genotype in phenotyping 1. \jri{shouldn't we expect a negative correlation betweeh Hop and TI on day 40? need to have an A and B in the figure and explain one is for pheno1 and one is for pheno2. please explain whiskers and dots on figure too.} \lev{sure, I mean, presumably we would expect things with Hop to have a smaller TI} \jri{yup, but we should mention that the expectation is negative/}

>>>>

%TO DO
%Fix figures and tables
%check citations in last paragraph for ter1
%reformat for AJB
%preprint somewhere
%finish fixing paragraph that jeff doesn't like
%check map details
%make figs smaller so they don't go to end of document
%Code for SAS and command line stuff on Git
%Phenotyping and genotyping data on Git (it's on Figshare?)

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE).  To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the \emph{Hopscotch} insertion may be adaptive under certain environmental conditions, we looked for significant associations between the \emph{Hopscotch} frequency and environmental variables using BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (\emph{i.e.,} Bayes Factor; \citealt{Coop2010}). We used genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional SNP (Table~S3). Environmental data were obtained from \url{www.worldclim.org}, the Harmonized World Soil Database \citep{FAOHWSD} and \url{www.harvestchoice.org} and summarized by principle component analysis \citep{Pyhajarvi2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \citep{Thornton2003} to calculate pairwise F\textsubscript{ST} between populations and to calculate standard diversity statistics (number of haplotypes, haplotype diversity, Watterson's estimator $\hat\theta_W$, pairwise nucleotide diversity $\hat\theta_\pi$, and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by \emph{Hopscotch} genotype we used Phylip (\url{http://evolution.genetics.washington.edu/phylip.html}) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the \emph{Hopscotch} insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \citealt{Chia2012}).

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and the same phased 55K SNP data from \citep{Pyhajarvi2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map \citep{Gerke2013}. We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

\subsection*{Phenotyping of \emph{parviglumis}}

To investigate the phenotypic effects of the \emph{Hopscotch} insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the \emph{Hopscotch} is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the \emph{Hopscotch} in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Following germination, 206 successful germinations were then planted into one-gallon pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day by hand and with an automatic drip containing 10-20-10 fertilizer.

Starting on day 15, we measured tillering index as the ratio of the sum of tiller lengths to the height of the plant \citep{Briggs2007}. Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index or variation at \emph{tb1}. Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (\url{http://www.ars.usda.gov}). We genotyped individuals for the \emph{Hopscotch} insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of Phenotyping 1, indicating that a minimum of 71 individuals in each genotypic class are needed to detect the observed effect of the \emph{Hopscotch} on tillering index.

We performed a second phenotyping experiment (Phenotyping 2) in which we germinated 372 seeds of \emph{parviglumis}, choosing equally between sites previously determined to have or not have the \emph{Hopscotch} insertion. Seeds were germinated and planted on day 7 post fruit-case removal into two gallon pots. Plants were watered twice daily, alternating between fertilized and non-fertilized water. We began phenotyping successful germinations (302) for tillering index on day 15 post fruit-case removal, and phenotyped every five days until day 50. At day 50 we measured culm diameter between the third and fourth nodes. We extracted DNA and genotyped plants following the same guidelines as in Phenotyping 1.

Resulting tillering index data for each genotypic class did not meet the criteria for a repeated measures ANOVA, so we transformed the data using a Box-Cox transformation ($\alpha=0$) \mbh{what is the alpha value here?} implemented in the car package in R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between \emph{Hopscotch} genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS.

@@

To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion from our initial genotyping. We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. We found the \emph{Hopscotch} segregating at a frequency of 0.65 with no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find a weak positive correlation between presence of the \emph{Hopscotch} and tillering index on day 40 (p=0.0848), but no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and \emph{Hopscotch} genotype, or culm diameter and \emph{Hopscotch} genotype in Phenotyping 1. \jri{shouldn't we expect a negative correlation betweeh Hop and TI on day 40? need to have an A and B in the figure and explain one is for pheno1 and one is for pheno2. please explain whiskers and dots on figure too.} \lev{sure, I mean, presumably we would expect things with Hop to have a smaller TI} \jri{yup, but we should mention that the expectation is negative/}
